Cluster Analysis

### 2020-07-31 MSDS 411 assign 4.Rmd
moved datafiles into the R project, updated code so it no longer references locally stored datafiles

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
employment <- read.csv(file="file:///Users/nadianoui-mehidi/Desktop/EuropeanEmployment.csv",head=TRUE,sep=",")
recidivism <- read.csv(file="file:///Users/nadianoui-mehidi/Desktop/recidivism.csv",head=TRUE,sep=",")
states <- read.csv(file="file:///Users/nadianoui-mehidi/Desktop/USStates.csv",head=TRUE,sep=",")
library(NbClust)
library(dplyr)
library(skimr)
library(cluster)
require(tidyverse)
library(kohonen)
require(ggplot2)
require(ggridges)
require(RColorBrewer)
library(vegan)
require(useful)
require(Hmisc)
library(HSAUR)
library(MVA)
library(HSAUR2)
library(fpc)
library(mclust)
library(lattice)
library(car)
library(cluster)
require(maptree)
library(factoextra)
library(data.table)
library(formattable)
library(tidyr)
```

@@

```{r cars}
print(employment)
```

##(1)	Initial Exploratory Data Analysis:  
When we have multivariate data, the scatterplot of each pair of variables can be used as the basis of an initial examination of the the data for informal evidence of cluster structure. Using the pairs plot below to scan the individual 2-dimensional views of the data we see that some industries move together. Agriculture has negative correlation with Service, Construction, and Manufacturing industries. Transport, manufacturing and construction are also correlated

@@

##(2)	Visualizing the Data with Labelled Scatterplots:  
While the pairs plot allows us to scan all pairwise scatterplots, it is not the ideal visualization of the data.  After we have honed in on some interesting dimensions we can create more specialized plots for those dimensions to see whether they reveal any clusters.

@@

##(4)	Hierarchical Clustering Analysis

@@

##(5)	k-Means Clustering Analysis  
The k-means algorithm is another method to cluster data. This technique partitions the data set into unique homogeneous clusters whose observations are similar to each other but different than other clusters.

@@

## 6.	Optimal Number of Clusters

@@

# 7. USSTATES: Hierarchical Cluster Analysis
The USSTATES dataset is a 12 variable dataset with 50 records representing the US States. Each row includes a group variable for region that we can use as our framework for understanding how states could be grouped. There are 2 continuous variables: population and household income. The remaining variables are state-wide average or proportion scores for various demographic  variables.  As such, higher scores for the composite variables translate into having more of that quality.

```{r}
print(states)
```

@@

# 8. RECIDIVISM: PAM Cluster Analysis  
The recidivism dataset is an 18 variable dataset with 1445 records.  The data is a random sample of convicts released from prison between July 1, 1977 and June 30, 1978. The information was collected retrospectively by looking at records in April 1984, so the maximum possible length of observation is 81 months.

```{r}
head(recidivism)
```

@@

### Distance Measure
In order for a clustering algorithm to yield sensible results, we have to use a distance metric that can handle mixed data types. In this case, we will use something called Gower distance. For each variable type, the distance metric that works well for that type is used and scaled to fall between 0 and 1. Our quantitative variables use a range-normalized Manhattan distance. Our binary variables use a Dice coefficient.
```{r}
r<-recidivism[,-c(18)]
dist <- daisy(r,
                    metric = "gower",
                    type = list(logratio = 3))
summary(dist)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(wooldridge)
data('recid')

#states <- read.csv(file="file:///Users/nadianoui-mehidi/Desktop/USStates.csv",head=TRUE,sep=",")
library(NbClust)
library(dplyr)
library(skimr)
library(cluster)
require(tidyverse)
library(kohonen)
require(ggplot2)
require(ggridges)
require(RColorBrewer)
library(vegan)
require(useful)
require(Hmisc)
library(HSAUR)
library(MVA)
library(HSAUR2)
library(fpc)
library(mclust)
library(lattice)
library(car)
library(cluster)
require(maptree)
library(factoextra)
library(data.table)
library(formattable)
library(tidyr)
```

@@

```{r}
#"/Users/nadianoui-mehidi/Cluster-Analysis"
#dir.create("raw_data")
#file.copy("~/Documents/EuropeanEmployment.csv", "raw_data")

```{r cars}

library(readr)
employment <- read.csv("raw_data/EuropeanEmployment.csv")
print(employment)

## Initial Exploratory Data Analysis:  
When we have multivariate data, the scatterplot of each pair of variables can be used as the basis of an initial examination of the the data for informal evidence of cluster structure. Using the pairs plot below to scan the individual 2-dimensional views of the data we see that some industries move together. Agriculture has negative correlation with Service, Construction, and Manufacturing industries. Transport, manufacturing and construction are also correlated

@@

## Visualizing the Data with Labelled Scatterplots:  
While the pairs plot allows us to scan all pairwise scatterplots, it is not the ideal visualization of the data.  After we have honed in on some interesting dimensions we can create more specialized plots for those dimensions to see whether they reveal any clusters.

@@

## Hierarchical Clustering Analysis

@@

## K-Means Clustering Analysis  
The k-means algorithm is another method to cluster data. This technique partitions the data set into unique homogeneous clusters whose observations are similar to each other but different than other clusters.

@@

## Optimal Number of Clusters

@@

#  USSTATES: Hierarchical Cluster Analysis
The USSTATES dataset is a 12 variable dataset with 50 records representing the US States. Each row includes a group variable for region that we can use as our framework for understanding how states could be grouped. There are 2 continuous variables: population and household income. The remaining variables are state-wide average or proportion scores for various demographic  variables.  As such, higher scores for the composite variables translate into having more of that quality.

```{r}
file.copy("~/Documents/USStates.csv", "raw_data")

states <- read.csv("raw_data/USStates.csv")

print(states)
```

@@

# RECIDIVISM: PAM Cluster Analysis  
The recidivism dataset is an 18 variable dataset with 1445 records.  The data is a random sample of convicts released from prison between July 1, 1977 and June 30, 1978. The information was collected retrospectively by looking at records in April 1984, so the maximum possible length of observation is 81 months.

```{r}
head(recid)
```

@@

### Distance Measure
In order for a clustering algorithm to yield sensible results, we have to use a distance metric that can handle mixed data types. In this case, we will use something called Gower distance. For each variable type, the distance metric that works well for that type is used and scaled to fall between 0 and 1. Our quantitative variables use a range-normalized Manhattan distance. Our binary variables use a Dice coefficient.
```{r}
r<-recid[,-c(18)]
dist <- daisy(r,
                    metric = "gower",
                    type = list(logratio = 3))
summary(dist)
```
